Transmembrane helices are also an overlooked source of major histocompatibility complex class II epitopes and evolutionary more conserved than expected by chance

### 2020-04-14 article.tex
Processed feedback Frans

% Affiliations
\usepackage{authblk}
\title{
  Transmembrane helices are also 
  an overlooked source of major histocompatibility complex class II epitopes
  and evolutionary more conserved than expected by chance
}
\author[1]{Richèl J.C. Bilderbeek}
\author[2]{Frans Bianchi}
\affil[1]{Groningen Institute for Evolutionary Life Sciences, University of 
Groningen, Groningen, The Netherlands}
\affil[2]{Frans' Institute, University of Groningen, Groningen, The Netherlands}

@@

Transmembrane helices (TMHs) in the human proteome
are an overrepresented source of epitopes on major 
histocompatibility complex (MHC) class I for the majority of HLA-I haplotypes. 
It is unknown if this same overrepresentation
is also the case for the proteome of a pathogen.
Additionally, it is unknown if TMH-II is also likelier to present
TMHs than expected by chance. 
Lastly, if the immune system if more tailored to detect polypeptides
derived from TMHs, can we confirm this by detecting the evolutionary
conservation of this behavior?
This study shows that MHC-I [also has/does not have] more
epitopes derived from a TMH for a pathogen proteome, when compared with
a host proteome.
Additionally, MHC-II binds to polypeptides derived from TMHs 
[less/equally/more] often than expected by chance.
Lastly, this study shows that in the mutation rate in TMHs of a pathogen
is [lower/as much/higher], hinting at 
evolutionary [conservation/neutrality/disruptive selection].
Our findings suggest that the immune system is [less/neutral/more]
vigilant to TMHs than expected by chance and this has left [a/no]
signal in the evolutionary history of the pathogen.

@@

For MHC-I, it was found that epitopes derived 
from human transmembrane helices (TMHs)
are over-presented by all human leukocyte antigen (HLA)-A and 
most HLA-B super types (\cite{bianchi2017}). 
One explanation is that the presentation of TMHs 
may have an evolutionary advantage for 
the (human) host, as TMHs may be more 
conserved due to the functional requirement of being able to span a lipid 
bilayer. 
Due to this, pathogens have a lower chance to develop an escape mutation,
as many mutations will result in a disfunctional TMH.

@@

If presentation of TMHs on MHC-I would bring an evolutionary advantage 
in the recognition of pathogens by the immune, 
it would follow that this is equally important for MHC-II.

@@

If the human immune system is more prone to detect polypeptides
derived from a pathogen's TMHs, one may expect to find an evolutionary
signal in fast-evolving pathogen. 
We can assume that only after the moment that COVID-19 spilled over 
to humans, it's the human immune system that is an influential component 
of its evolutionary selection.
Using the extensive amount of time-dated COVID-19 sequences,
we can quantify to which extent selection in COVID-19 occurs
in the transmembrane part of its proteins.

@@

Bianchi and colleagues obtained a distribution of 
percentages of MHC-I epitopes overlap with TMHs in Homo sapiens
for the different HLA haplotypes, with has an average of 5.3\%.
We obtained a similar distribution of percentages of MHC-I epitopes that 
overlap with TMHs for the different HLA haplotypes, but then applied to
COVID-19.

We compare these distribution using a two-sample 
Kolmogorov-Smirnov (KS) test. The KS test determines if two samples
are derived from a same distribution, without making assumptions
regarding the shape of that distribution.

@@

\iffalse
	Transmembrane helices and strong MHC-II-binding peptides
	were predicted for a tuberculosis reference proteome 
	from \url{https://www.ebi.ac.uk/reference_proteomes}
	using the \verb;epitopeome; R package (\cite{epitopeome}).
	We picked the three MHC-II alleles that are most abundant 
	in the current human population, 
	which are DRB4*0101, HLA-DPA10103-DPB10402 
	\richel{
	  We decided for DPB1*0402, but there are multiple alleles matching.
	  I just picked the first one of the alleles matching, which are:
	  HLA-DPA10x-DPB10402, where x equals 103-110, 201-204, 301-303 or 401
	}
	and DQA1*0501/DQB1*0301 \richel{reference here}.
	The 5\% peptides with the lowest IC50 values were defined as binders.
	We then simply counted the number of amino acids that were present inside the
	cell, within the membrane or outside of the cell, as well as if it was part 
	of a strong MHC-II binding site, as shown in figure \ref{tab:results}.

@@

\iffalse
	The evolutionary conservation of TMHs was measured from a
	DNA alignment of multiple transmembrane proteins in multiple
	species of the mycoplasma bacterial family
	\richel{how obtained exactly? how alignment done?}.

	Using the \verb;tmhprot; R package (\cite{tmhprot}), the DNA alignment is split
	into two alignments, one for the TMH parts, another for the non-TMH
	parts. Each alignment was tested by the \verb;mcbette; 
	R package (\cite{mcbette}) to select the Bayesian inference model with
	the highest evidence (a.k.a. the marginal likelihood) using the nested
	sampling approach as described in \cite{maturana2018model},
	using the popular Bayesian phylogenetic tool 
	BEAST2 (\cite{bouckaert2014beast}) in the back-end.
	\verb;mcbette; used a set of 40 candidate inference models, 
	consisting of all combinations of 
	4 site models (JC, HKY, TN, GTR), 
	2 clock models (strict, RLN) and 
	5 tree priors (Yule, BD, CBS, CCP, CEP).

@@

Figure \ref{fig:1} shows the percentages of epitopes overlapping 
with TMHs for a human and COVID-19 proteome.
The KS test to determine if these percentages are sampled from
a same distribution resulted in a p-value of \richel{unknown},
which makes us reject/accept the hypothesis that these percentages
are sampled from a same distribution.

\begin{figure}[!htbp]
  \includegraphics[width=\textwidth]{bbbq_1/fig_bbbq_1_watermarked.png}
  \caption{
    \% epitopes overlapping with transmembrane helix
    for a human and COVID-19 proteome
  }
  \label{fig:1}
\end{figure}

@@

\iffalse
	We compared the mutation rates between the TMH and non-TMH part of
	multiple mycoplasma species. Where we expect no variation 
	in mutation rate for every TMH amino acid,
	\richel{
	  we can test this, but unsure if that would make sense
	}
	we know that non-TMH part will have regions of different evolutionary
	conservation: functional domains, especially in protein-protein
	interactions will be strongly conserved, due to an even more constrained
	set of peptides that enable a certain function.

>>>>

% Affiliations
\usepackage{authblk}
\title{
  Transmembrane helices are also 
  an overlooked source of major histocompatibility complex class II epitopes
  and evolutionary more conserved than expected by chance
}
\author[1]{Richèl J.C. Bilderbeek}
\author[1]{Maxim Baranov}
\author[1]{Geert van den Bogaart}
\author[1]{Frans Bianchi}
\affil[1]{GBB, University of 
Groningen, Groningen, The Netherlands}

@@

Transmembrane helices (TMHs) in the human proteome
are an overrepresented potential source of epitopes on major 
histocompatibility complex (MHC) class I for the majority of HLA-I haplotypes. 
It is unknown if this same overrepresentation
also holds for the proteome of a pathogen.
Additionally, it is unknown if TMH-II is also likelier to present
TMHs than expected by chance. 
Lastly, if the immune system if more tailored to detect polypeptides
derived from TMHs, can we confirm this by detecting the evolutionary
conservation of this behavior?
This study shows that MHC-I [also has/does not have] more
epitopes derived from a TMH for a pathogen proteome, when compared with
a host proteome.
Additionally, MHC-II binds to polypeptides derived from TMHs 
[less/equally/more] often than expected by chance.
Lastly, this study shows that in the mutation rate in TMHs of a pathogen
is [lower/as much/higher], hinting at 
evolutionary [conservation/neutrality/disruptive selection].
Our findings suggest that the immune system is [less/neutral/more]
vigilant to TMHs than expected by chance and this has left [a/no]
signal in the evolutionary history of the pathogen.

@@

For MHC-I, it was found that predicted epitopes derived 
from human transmembrane helices (TMHs)
are over-presented by all human leukocyte antigen (HLA)-A and 
most HLA-B super types (\cite{bianchi2017}). 
One explanation is that the presentation of TMHs 
may have an evolutionary advantage for 
the (human) host, as TMHs have a reduced variability 
due to the functional requirement of being able to span a lipid 
bilayer. 
Due to this, pathogens have a lower chance to develop an escape mutation,
as many mutations will result in a disfunctional TMH.

@@

If presentation of TMHs on MHC-I would bring an evolutionary advantage 
in the recognition of pathogens by the immune system, 
it would follow that this is equally important for MHC-II, 
especially as CD4+ T cell help is needed for a long lasting CD8+Tcell 
response \cite{novy2007cd4}.

@@

If the human immune system is more prone to detect polypeptides
derived from a pathogen's TMHs, one may expect to find an evolutionary
signal in fast-evolving pathogen. 
We can assume that only after the moment that COVID-19 spilled over 
to humans, it's the human immune system that is an influential component 
of its evolutionary selection.
Using the extensive amount of time-dated COVID-19 sequences,
we can quantify to which extent selection in COVID-19 occurs
in the transmembrane part of its proteome.

@@

Bianchi and colleagues obtained a distribution of 
percentages of MHC-I epitopes overlap with TMHs in Homo sapiens
for the different HLA haplotypes, with an average of 5.3\%.
We obtained a similar distribution of percentages of MHC-I epitopes that 
overlap with TMHs for the different HLA haplotypes, but then applied to
COVID-19.

We compare these distributions using a two-sample 
Kolmogorov-Smirnov (KS) test. The KS test determines if two samples
are derived from the same distribution, without making assumptions
regarding the shape of that distribution.

@@

\iffalse
	Transmembrane helices and strong MHC-II-binding peptides
	were predicted for a tuberculosis reference proteome 
	from \url{https://www.ebi.ac.uk/reference_proteomes}
	using the \verb;epitopeome; R package (\cite{epitopeome}).
	We picked the three MHC-II alleles that are most abundant 
	in the current human population, 
	which are DRB4*0101, HLA-DPA10103-DPB10402 
	\richel{
	  We decided for DPB1*0402, but there are multiple alleles matching.
	  I just picked the first one of the alleles matching, which are:
	  HLA-DPA10x-DPB10402, where x equals 103-110, 201-204, 301-303 or 401
	}
	and DQA1*0501/DQB1*0301 \richel{reference here}
	\frans{
	  Neem aan dat dit die ref is naar dat stukje van most common haplotype in human population. Misschien moeten we even de coverage erbij melden. 
	  Greenbaum, Jason, et al. "Functional classification of class II human leukocyte antigen (HLA) molecules reveals seven different supertypes and a surprising degree of repertoire sharing across supertypes." Immunogenetics 63.6 (2011): 325-335.
	}.
	The 5\% peptides with the lowest IC50 values were defined as binders.
	We then simply counted the number of amino acids that were present inside the
	cell, within the membrane or outside of the cell, as well as if it was part 
	of a strong MHC-II binding site, as shown in figure \ref{tab:results}.

@@

\iffalse

	\frans{
	  Ik heb zitten denken over de evolutie naast conservering is occurrence ook heel belangrijk. Moeten we de mate van conservering niet wegen ten opzichte van de occurrence.
	  Wat ik daarmee bedoel is data als je kiest voor een mega geconseveerd epitope maar van dat type komen er maar 40 in een proteome voor versus een minder geconserveerd epitope maar daar zijn wel 100.000 varianten van dan heb je meer kans om 1 van die 100.000 te pakken toch? 
	}
	The evolutionary conservation of TMHs was measured from a
	DNA alignment of multiple transmembrane proteins in multiple
	species of the mycoplasma bacterial family
	\richel{how obtained exactly? how alignment done?}
	\frans{
	Via entrez en blast 
	  Ik he been transcript van Maxim ik zal dit proberen te vertalen naar een stukje tekst. 
	}
	.
	Using the \verb;tmhprot; R package (\cite{tmhprot}), the DNA alignment is split
	into two alignments, one for the TMH parts, another for the non-TMH
	parts. Each alignment was tested by the \verb;mcbette; 
	R package (\cite{mcbette}) to select the Bayesian inference model with
	the highest evidence (a.k.a. the marginal likelihood) using the nested
	sampling approach as described in \cite{maturana2018model},
	using the popular Bayesian phylogenetic tool 
	BEAST2 (\cite{bouckaert2014beast}) in the back-end.
	\verb;mcbette; used a set of 40 candidate inference models, 
	consisting of all combinations of 
	4 site models (JC, HKY, TN, GTR), 
	2 clock models (strict, RLN) and 
	5 tree priors (Yule, BD, CBS, CCP, CEP).

@@

Figure \ref{fig:1} shows the percentages of epitopes overlapping 
with TMHs for a human and COVID-19 proteome.
The KS test to determine if these percentages are sampled from
a same distribution resulted in a p-value of \richel{unknown},
which makes us reject/accept the hypothesis that these percentages
are sampled from the same distribution.

\begin{figure}[!htbp]
  \includegraphics[width=\textwidth]{bbbq_1/fig_bbbq_1_watermarked.png}
  \caption{
    \% epitopes overlapping with transmembrane helix
    for a human and COVID-19 proteome.
    \richel{
      The Illuminati watermark indicates the data underlying this
      figure has been simulated
    }
  }
  \label{fig:1}
\end{figure}

@@

\iffalse
	\frans{
	  De TMH evolutie data van COVID is missachien interresanter. Ik zie het als volgt. 
	  COVID is een mooi voorbeeld van een virus 
	  Myco van een bacterie
	  In de bacterie is duidelijk dat tenmisnte voor MHCII TMHs zijn overgrepresenteerd virus niet. 
	  Dus verhaal kan zijn dat TMHs belangrijk voot bacteirele afweer niet zozeer voor virussen.

	  Hiervoor moeten we MHC-I doen voor myco
	  Kijken wat evolutie data doet voor virus dan evt kijken wat dit doet voor myco en vergelijken. 
	}

	We compared the mutation rates between the TMH and non-TMH part of
	multiple mycoplasma species. Where we expect no variation 
	in mutation rate for every TMH amino acid,
	\richel{
	  we can test this, but unsure if that would make sense
	}
	we know that non-TMH part will have regions of different evolutionary
	conservation: functional domains, especially in protein-protein
	interactions will be strongly conserved, due to an even more constrained
	set of peptides that enable a certain function.
